### 2020-03-09 manuscript/manuscript.rmd
update manuscirpt with sensitive analyses

---
bibliography: manuscript.bib
csl: nature.csl
output:
  github_document: default
  pdf_document: default
  word_document: default
  html_document: default
---

@@

# Rooting the animal tree of life

@@

## Abstract
Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool phylobayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary Results. Briefly, the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand). The rate of excahnge can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites. The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary. Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software tool CAT and has been widely applied to questions of deep animal phylogeny, and the C models of iqtree that have until now not been applied to the problem.

@@

Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig. 1C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated-- Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.

@@

Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving]. Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie]. Feuda *et al*. [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six]. Thus, it is important to reevalute the performance of the impact of data-recoding compared to non-recoding methods in animal phylogeny.

## Overview of published analyses
Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies (Table 1). These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Extended Data Fig. 1). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Extended Data Fig. 1). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 4; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C10-C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and site-homogeneous models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models. Thus, our reanalyzing results are largely consistent with previous analyses, both suggesting that the support of Porifera-sister is increasing with more complex models and closely related outgroups. More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig. 5). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in Whelan2017_full matrices.

@@

@@

## The current state of understanding and future direction
Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices. By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases. Three factors contribute significantly to the previous incongruent results based on our analyses. First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies. Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models. More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice. Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species. Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister. Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.

@@

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  html_document: default
  pdf_document: default
  word_document: default
  github_document: default
---

@@

# Rooting the animal tree of life - more is less for Ctenophora-sister?

@@

## Abstract (ideally less than 200 words)

Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships in the animal tree of life. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions. We next quantify the  support of phylogenetic signal and conduct sensitive analyses in several representative datasets. We found that the Porifera-sister can only be recovered with a combined factors in certain data matrices, with inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories and more closely related outgroups, whereas Ctenophora-sister is strongly supported in most other cases. We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic. Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary Results. Briefly, the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites. The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary. Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software PhyloBayes  and has been widely applied to questions of deep animal phylogeny, and the CAT-like C10 – C60 models (C models) employed in maximum likelihood (ML) inference until now has not been applied to the problem.

@@

Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig. 1C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated-- Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.

@@

Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving]. Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie]. Feuda *et al*. 2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan *et al*. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six]. Thus, it is important to reevalute the performance of the impact of data-recoding compared to non-recoding methods in animal phylogeny.

## Overview of published analyses
Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 1 data matrices from 12 studies (Table 1). These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig. 1). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Extended Data Fig. 1). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in all other cases.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 4; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive. The GTR+C60 models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous+C60 models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C60 models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared IQtree C60 results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models. These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups. More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Fig. 5). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models. This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C60 models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Whelan2017_full matrices.

@@

### Sensitive test of CAT categories {place holder for sensitive results}
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix. Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx). Although the true number of substitutional categories in empirical data matrices is impossible with current bioinformatic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results,  potentially indicating that the C60 models might be already overfitting to most matrices. These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses.

We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices. Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices.Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories. The major transition occurred at 90 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four choanimalia matrices. Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than site-homogeneous models.

The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model. Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups. Interestingly, we found both Whelan2017_full and Whelan2017_strict matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.  
Thus, our results also strongly suggested that the exchangeability matrix of evolutionary model (a more simplistic Poisson model and more realistic GTR model) also has strong impact to the rooting position of animal phylogeny in PhyloBayes.

@@

## The current state of understanding and future direction
Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices. By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases. Four factors contribute significantly to the previous incongruent results based on our analyses. First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies. Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models. More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice. Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny. Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species. Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister. Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.

@@

#### Sensitive analyses with different number of substitutional categories
